---
title: EHS metrics - HHRIP
subtitle:
author:
  - name: Eliot Monaco
    affiliation:
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_notebook
---

# Purpose

This is an analysis of Healthy Homes Rental Inspection Program inspections data set. Data quality is checked, and mean days to resolution is calculated for complaints.

Data for this analysis is from the Inspection History Table. Each row represents one inspection. Complaints may involve more than one inspection, therefore the table is pivoted so that each row represents one complaint. Only complaints that begin on or after 2023-01-01 are summarized. The start date for a complaint is the date of the first "Complaint" inspection. The end date is the date of the last "Admin Closure" or "Desk Approval" inspection.

# Set up environment and import data

```{r setup}
knitr::opts_chunk$set(
  fig.width = 9,
  fig.height = 4
)
```

```{r message=FALSE}
library(tidyverse)
library(readxl)
library(setmeup)
library(openxlsx2)
```

Import inspections data.

```{r}
insp1 <- read_excel(
  "../data/1-source/InspectionHistoryTable.xlsx",
  .name_repair = ~ fix_colnames(.x, n_pfx = "var_")
)
```

# Prep data

Convert `inspection_date` from POSIXct to Date class.

```{r}
insp1$inspection_date <- as.Date(insp1$inspection_date)
```

## Validation

```{r}
dq_review <- list()
```

Count missing values.

```{r}
vars <- c(
  "id",
  "complaint_id",
  "inspection_type",
  "inspection_date",
  "property_zip_code"
)

data.frame(
  variable = vars,
  n_missing = unlist(lapply(vars, \(x) sum(is.na(insp1[[x]]))))
)
```

```{r}
# Data set earliest possible date
dstart <- as.Date("2018-08-01")
# Data set latest possible date
dexport <- as.Date("2025-03-26")
# Data set start date for metric calculations
dmetrics <- as.Date("2023-01-01")
```

Rows missing a `complaint_id` pre- & post-`r as.character(dmetrics)`.

```{r}
data.frame(
  pre = nrow(insp1[is.na(insp1$complaint_id) & insp1$inspection_date < dmetrics, ]),
  post = nrow(insp1[is.na(insp1$complaint_id) & insp1$inspection_date >= dmetrics, ])
)
```

Non-numeric `id`s.

```{r}
insp1 |>
  filter(grepl("\\D", id)) |>
  arrange(id) |>
  distinct(id)
```

Duplicate `id`s.

```{r}
any(duplicated(insp1$id))
```

Non-numeric `complaint_id`s.

```{r}
insp1 |>
  filter(grepl("\\D", complaint_id)) |>
  filter(!grepl("^\\d+[\\.-]\\d+$", complaint_id)) |>
  arrange(complaint_id) |>
  distinct(complaint_id)
```

```{r}
dq_review$problem_complaint_id <- insp1 |>
  filter(
    is.na(complaint_id) |
      (grepl("\\D", complaint_id) & !grepl("^\\d+[\\.-]\\d+$", complaint_id))
  ) |>
  select(id, complaint_id, inspection_type, inspection_date) |>
  arrange(complaint_id)
```

Invalid `inspection_date`s (dates occurring before program start date or after data export).

```{r}
dq_review$invalid_date <- insp1 |>
  filter(
    inspection_date < dstart |
      inspection_date > dexport
  ) |>
  select(id, complaint_id, inspection_type, inspection_date) |>
  arrange(complaint_id)

insp1 |>
  filter(
    inspection_date < dstart |
      inspection_date > dexport
  ) |>
  arrange(inspection_date) |>
  select(inspection_date)
```

Validate `inspection_type` values.

```{r}
insp1 |>
  group_by(inspection_type) |>
  count()
```

Invalid zip codes.

```{r}
insp1 |>
  filter(!grepl("^\\d{5}$", property_zip_code)) |>
  arrange(property_zip_code) |>
  distinct(property_zip_code)
```

Clean zip codes.

```{r}
insp1 <- insp1 |>
  mutate(property_zip_code = if_else(
    grepl("\\d{5}", property_zip_code),
    str_extract(property_zip_code, "\\d{5}"),
    NA
  ))
```

`complaint_id`s associated with > 1 zip code.

```{r}
insp1 |>
  group_by(complaint_id, property_zip_code) |>
  count() |>
  ungroup() |>
  find_dupes("complaint_id")
```

## Config

- Select relevant variables.
- Remove rows with missing or invalid data that cannot be included in the analysis.
- Remove rows with dates after the data was exported.
- Sort by `inspection_date`.
- Assign ID to unique combinations of `complaint_id`, `inspection_type`, and `inspection_date`.

```{r}
insp2 <- insp1 |>
  select(all_of(vars)) |>
  filter(!if_any(c(complaint_id, inspection_type, inspection_date), is.na)) |>
  filter(!inspection_date > dexport) |>
  arrange(inspection_date) |>
  assign_id(
    vars = c("complaint_id", "inspection_type", "inspection_date"),
    id_name = "new_id",
    prefix = "n"
  )
```

Deduplicate on `new_id`.

```{r}
insp2 <- insp2 |>
  distinct(new_id, .keep_all = TRUE)
```

Create a variable to number rows that have the same `complaint_id` and `inspection_type`.

```{r}
insp2$n <- ave(
  x = insp2$id,
  insp2$complaint_id,
  insp2$inspection_type,
  FUN = function(x) sprintf("%02d", seq_along(x))
)
```

Combine `inspection_type` and `n` into new variable.

```{r}
insp2 <- insp2 |>
  unite(
    col = "inspection_type_n",
    inspection_type, n,
    sep = " ",
    remove = TRUE,
    na.rm = TRUE
  )
```

Convert `inspection_type_n` to factor.

```{r}
vars <- unique(insp2$inspection_type_n)
insp_types <- c("complaint", "reinspection", "field", "desk", "admin")
p <- paste0("(?i)^", insp_types)
lvl <- unlist(lapply(p, \(x) sort(vars[grepl(x, vars)])))

insp2 <- insp2 |>
  mutate(inspection_type_n = factor(inspection_type_n, levels = lvl))
```

```{r}
insp2 |>
  group_by(inspection_type_n) |>
  count()
```

Pivot `insp2` to one row per complaint, one column per zip code.

```{r}
zips <- insp2 |>
  select(id, complaint_id, property_zip_code) |>
  distinct(complaint_id, property_zip_code, .keep_all = TRUE) |>
  filter(!is.na(property_zip_code))

zips$n <- ave(
  x = zips$id,
  zips$complaint_id,
  FUN = seq_along
)

zips <- zips |>
  select(-id) |>
  mutate(zip = "zip") |>
  pivot_wider(
    names_from = c(zip, n),
    values_from = property_zip_code
  )
```

Pivot `insp2` to one row per complaint.

```{r}
comp1 <- insp2 |>
  select(complaint_id, inspection_type_n, inspection_date) |>
  pivot_wider(
    names_from = inspection_type_n,
    values_from = inspection_date,
    names_sort = TRUE,
    names_repair = fix_colnames
  ) |>
  arrange(complaint_01)
```

Join zip codes.

```{r}
comp1 <- comp1 |>
  left_join(zips, by = "complaint_id")
```

Rows with multiple zip codes.

```{r}
dq_review$multiple_zip <- comp1 |>
  filter(!is.na(zip_2))

comp1 |>
  filter(!is.na(zip_2)) |>
  select(complaint_id, starts_with("zip"))
```

Rows with multiple zip codes pre- & post-`r as.character(dmetrics)`.

```{r}
data.frame(
  pre = nrow(comp1[!is.na(comp1$zip_2) & comp1$complaint_01 < dmetrics, ]),
  post = nrow(comp1[!is.na(comp1$zip_2) & comp1$complaint_01 >= dmetrics, ])
)
```

Rows missing an initial (complaint) inspection.

```{r}
dq_review$missing_start_date <- comp1 |>
  filter(is.na(complaint_01))

comp1 |>
  filter(is.na(complaint_01))
```

Rows missing a final (admin closure or desk approval) inspection.

```{r}
dq_review$missing_end_date <- comp1 |>
  filter(if_all(matches("^desk|^admin"), is.na))

comp1 |>
  filter(if_all(matches("^desk|^admin"), is.na))
```

## Check dates

Check order of inspection dates.

- Find minimum non-complaint inspection date.

```{r}
p <- paste0("^", insp_types[2:5], ".+\\d", collapse = "|")

comp1$alt_min_date <- apply(
  comp1[, grepl(p, colnames(comp1))],
  1,
  \(x) {
    ifelse(
      any(!is.na(x)),
      min(x, na.rm = TRUE),
      NA
    )
  }
) |>
  as.Date()
```

- Find maximum non-admin-closure/desk-approval inspection date.

```{r}
p <- paste0("^", insp_types[1:3], ".+\\d", collapse = "|")

comp1$alt_max_date <- apply(
  comp1[, grepl(p, colnames(comp1))],
  1,
  \(x) {
    ifelse(
      any(!is.na(x)),
      max(x, na.rm = TRUE),
      NA
    )
  }
) |>
  as.Date()
```

Rows where a non-complaint inspection precedes a complaint inspection.

```{r}
dq_review$late_start_date <- comp1 |>
  filter(complaint_01 > alt_min_date)

comp1 |>
  filter(complaint_01 > alt_min_date)
```

Rows where a non-admin-closure/desk-approval inspection follows an admin closure or desk approval inspection.

```{r}
dq_review$early_end_date <- comp1 |>
  filter(if_any(matches("^desk|^admin"), ~ .x < alt_max_date))

comp1 |>
  filter(if_any(matches("^desk|^admin"), ~ .x < alt_max_date))
```

Find the range between the minimum and maximum dates per complaint.

```{r}
p <- paste0("^", insp_types, ".+\\d", collapse = "|")

comp1$total_range <- apply(
  comp1[, grepl(p, colnames(comp1))],
  1,
  \(x) {
    ifelse(
      any(!is.na(x)),
      difftime(
        as.Date(max(x, na.rm = TRUE)),
        as.Date(min(x, na.rm = TRUE)),
        units = "days"
      ),
      NA
    )
  }
)
```

Plot ranges.

```{r}
comp1 |>
  ggplot(aes(x = total_range)) +
  geom_histogram(binwidth = 10)
```

Rows where `total_range` exceeds 1 year.

```{r}
comp1 |>
  filter(total_range > 365)
```

## Time to resolution

```{r}
# List to count complaints removed from summaries
rmvd <- list()

# Complaints removed due to missing "complaint" inspection date
rmvd$no_comp_date <- list(
  reason = "Missing inspection date for \"complaint\" inspection",
  n = comp1 |>
    filter(is.na(complaint_01)) |>
    nrow()
)
```

Filter by complaint inspection date on or after 2023-01-01.

```{r}
comp1 <- comp1 |>
  filter(!is.na(complaint_01), complaint_01 >= dmetrics)
```

Create `start_date` and `end_date` variables.

```{r}
comp1 <- comp1 |>
  mutate(start_date = complaint_01)

comp1$end_date <- apply(
  comp1[, grepl("^desk|^admin", colnames(comp1))],
  1,
  \(x) {
    ifelse(
      any(!is.na(x)),
      max(x, na.rm = TRUE),
      NA
    )
  }
) |>
  as.Date()
```

Validate date order of `start_date` and `end_date`.

```{r}
comp1 <- comp1 |>
  mutate(dates_sorted = is_sorted(start_date, end_date))

comp1 |>
  filter(!dates_sorted)
```

Remove rows with invalid date order.

```{r}
# Complaints removed due to mis-ordered start and end dates
rmvd$invalid_date_order <- list(
  reason = "End date precedes start date",
  n = comp1 |>
    filter(!dates_sorted) |>
    nrow()
)

comp1 <- comp1 |>
  filter(dates_sorted)
```

Calculate days to resolution.

```{r}
comp1 <- comp1 |>
  mutate(
    days_to_resolution = as.numeric(difftime(
      end_date,
      start_date,
      units = "days"
    )),
    month_started = format(start_date, "%Y-%m"),
    month_resolved = format(end_date, "%Y-%m")
  )
```

```{r}
# Complaints removed due to lack of end date
rmvd$no_end_date <- list(
  reason = "No desk approval or admin closure inspection date",
  n = comp1 |>
    filter(is.na(end_date)) |>
    nrow()
)

# Complaints removed due to multiple ZIP codes
rmvd$multiple_zips <- list(
  reason = "Multiple ZIP codes",
  n = comp1 |>
    filter(!is.na(zip_2)) |>
    nrow()
)
```

# Summarize

```{r}
complaint_sum <- list()
```

Summarize mean days to resolution by month.

```{r}
complaint_sum$days_to_res_mo <- comp1 |>
  filter(!is.na(days_to_resolution)) |>
  group_by(month_resolved) |>
  summarize(
    mean_days_to_resolution = round_ties_away(mean(days_to_resolution), 1),
    n_resolved = n()
  ) |>
  ungroup()
```

Summarize mean days to resolution by ZIP code.

```{r}
# complaint_sum$days_to_res_zip <- comp1 |>
#   filter(!is.na(days_to_resolution)) |>
#   filter(is.na(zip_2)) |>
#   group_by(zip_1) |>
#   summarize(
#     mean_days_to_resolution = round_ties_away(mean(days_to_resolution), 1),
#     n_resolved = n()
#   ) |>
#   ungroup()
```

Summarize resolved and unresolved complaints per month.

```{r}
complaint_sum$res_status <- comp1 |>
  group_by(month_started) |>
  summarize(
    n_total = n(),
    n_resolved = sum(!is.na(end_date)),
    n_unresolved = sum(is.na(end_date))
  ) |>
  mutate(
    pct_resolved = pct(n_resolved, n_total),
    pct_unresolved = pct(n_unresolved, n_total)
  ) |>
  select(
    month_started,
    n_resolved,
    pct_resolved,
    n_unresolved,
    pct_unresolved,
    n_total
  )
```

# Plot

Plot mean days to resolution.

```{r}
complaint_sum$days_to_res_mo |>
  ggplot(aes(x = month_resolved, y = mean_days_to_resolution)) +
  geom_col() +
  guides(x = guide_axis(angle = 45))
```

Plot resolution status.

```{r}
complaint_sum$res_status |>
  rename(resolved = pct_resolved, unresolved = pct_unresolved) |>
  pivot_longer(
    cols = c(resolved, unresolved),
    names_to = "resolution_status",
    values_to = "pct"
  ) |>
  ggplot(aes(x = month_started, y = pct, fill = resolution_status)) +
  geom_bar(stat = "identity") +
  scale_y_reverse() +
  guides(x = guide_axis(angle = 45))
```

# Data quality review

The following data quality issues were found. Records will be exported and provided to HHRIP for review.

```{r}
data.frame(
  reason = names(dq_review),
  total = unlist(lapply(dq_review, nrow)),
  since_2023 = unlist(lapply(
    dq_review,
    \(x) {
      if ("complaint_01" %in% colnames(x) && any(!is.na(x$complaint_01))) {
        x |>
          filter(complaint_01 >= dmetrics) |>
          nrow()
      } else {
        NA
      }
    }
  )),
  row.names = NULL
)
```

# Save and export

## Metrics summaries

```{r}
saveRDS(complaint_sum, "../data/3-final/complaint_summary_tables.rds")
```

```{r}
complaint_sum2 <- complaint_sum |>
  lapply(\(x) {
    colnames(x) <- colnames(x) |>
      gsub(pattern = "_", replacement = " ") |>
      str_replace(pattern = "^(n|pct)\\s(.+)$", replacement = "\\2 (\\1)") |>
      str_replace(pattern = "\\(pct\\)", replacement = "\\(%\\)") |>
      str_to_sentence()
    x
  })
```

```{r}
sheet_names <- c(
  "Days to resolution",
  "Resolution status"
)

wb <- wb_workbook()

for (i in 1:length(complaint_sum2)) {
  wb <- wb |>
    wb_add_worksheet(sheet = sheet_names[i]) |>
    wb_add_data(x = complaint_sum2[[i]])
}

wb_save(wb, "../output/hhrip-complaint-metric-summary.xlsx")
```

## Records to review

```{r}
saveRDS(dq_review, "../data/3-final/data_quality_review_tables.rds")
```

```{r}
ls <- "
The tables in this file contain rows that didn't pass certain quality checks
performed on the Inspection History Table.

- Problem complaint ID: complaint_id is missing or contains unexpected characters
- Invalid date: inspection_date is outside the expected date range for the data set
- Multiple ZIP: multiple ZIP codes associated with the same complaint_id
- Missing start date: complaint_id has no Complaint inspection
- Missing end date: complaint_id has no Desk Approval or Admin Closure inspection
- Late start date: Complaint inspection occurs after another inspection type
- Early end date: Desk Approval or Admin Closure inspection occurs before
  another inspection type
"

ls <- strsplit(ls, "\\n")
names(ls) <- "data_note"
dq_review <- c(ls, dq_review)
```

```{r}
sheet_names <- names(dq_review) |>
  gsub(pattern = "_", replacement = " ") |>
  str_to_sentence()

wb <- wb_workbook()

for (i in 1:length(dq_review)) {
  wb <- wb |>
    wb_add_worksheet(sheet = sheet_names[i]) |>
    wb_add_data(x = dq_review[[i]])
}

wb_save(wb, "../output/hhrip-data-quality-review.xlsx")
```
